\textbf{The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception}

### 2024-07-04 paper/manuscript.tex
knitted pdf

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants were presented with (fake) physiological feedback (the polygraph condition), instead of a face of a person they had to lie to (the interrogation condition), their response time for both lies and truths increased, as did their heart rate. Although the condition did not impact the subjective confidence that participants had in their answers, it suggests that believing oneself to be submitted to a machine that is supposedly able to detect deception by interpreting physiological signals is a harder and/or more stressful condition than lying to a person. While research linking interoception and deception is limited, our results are in line with studies that show an association between interoceptive awareness and anxiety (Domschke et al., 2010; Garfinkel \& Critchley, 2013; Yoris et al., 2015). Specifically, enhancing one's attention towards their internal bodily signals could have resulted in a hyper-vigilance towards physiological sensations that is perceived negatively. This is consistent with previous deception detection studies, in which participants' spontaneous lying behaviour only decreased when they were given feedback by a polygraph machine, but not when no feedback was given (Peleg et al., 2019). By extension, our study adds to the controversial discourse surrounding the use of physiological measures in past deception research, further questioning its validity as an indicator of deception (Oviatt et al., 2018; Rosky, 2013).

Our results suggest that higher ToM abilities were related to slower and less confident lies, but only in the polygraph condition. While previous bodies of work have reported mixed findings regarding the association between interoception and ToM (Chiou \& Lee, 2013; Gendolla \& Wicklund, 2009; Scaffidi Abbate et al., 2016; Wundrack \& Specht, 2023), our results suggest the two are negatively linked. One possible interpretation of our findings is that people with stronger ToM abilities by default rely more on their social skills and altercentric inference when lying (i.e., they focus on - and try to read - the other person). When that mechanism is unavailable or unsuited (e.g., when there is no person to lie to - but a ``machine'' in our case), their corresponding lying ability decreases. However, in light of the current field of mixed findings relating interoception and ToM (Canino et al., 2022; Gao et al., 2019; Miller, 2015; Shah et al., 2017), future studies are necessary to investigate the interaction of these mechanisms in different social contexts.

We also showed that interoceptive abilities (as indicated by the composite interoception scores) are correlated with a higher confidence in one's lies in the polygraph condition, a condition in which the attention towards internal reactions is fostered. Indeed, this is in line with previous studies that found individuals with low interoception were more averse to risk when reputational stakes were high, telling fewer egoistical lies (Vabba et al., 2022). In fact, Vabba et al. (2022) further reported that people with high interoception abilities were less likely to differ in risk-taking tendencies, telling the same number of lies regardless of the social stakes. Consistent with our results, Mohr et al. (2023) found that individuals with high interoceptive accuracy were more likely to make egocentric decisions. However, in contrast to previous studies (Füstös et al., 2013; Owens et al., 2018; Pinna \& Edwards, 2020; Pollatos et al., 2007), we did not find any significant relationship between individuals' interoception scores and their heart rate changes during their answers. This points toward a predominantly meta-cognitive effect without necessarily an actual bodily regulation (i.e., participants with good interoception feel that their lies are more convincing, but do not actively attenuate their bodily reactions).

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-domschke2010interoceptive}{}}%
Domschke, K., Stevens, S., Pfleiderer, B., \& Gerlach, A. L. (2010). Interoceptive sensitivity in anxiety and anxiety disorders: An overview and integration of neurobiological findings. \emph{Clinical Psychology Review}, \emph{30}(1), 1--11.

@@

\leavevmode\vadjust pre{\hypertarget{ref-garfinkel2013interoception}{}}%
Garfinkel, S. N., \& Critchley, H. D. (2013). Interoception, emotion and brain: New insights link internal physiology to social behaviour. Commentary on: {``Anterior insular cortex mediates bodily sensibility and social anxiety''} by terasawa et al.(2012). \emph{Social Cognitive and Affective Neuroscience}, \emph{8}(3), 231--234.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-peleg2019lie}{}}%
Peleg, D., Ayal, S., Ariely, D., \& Hochman, G. (2019). The lie deflator-the effect of polygraph test feedback on subsequent (dis) honesty. \emph{Judgment \& Decision Making}, \emph{16}(6).

@@

\leavevmode\vadjust pre{\hypertarget{ref-yoris2015roles}{}}%
Yoris, A., Esteves, S., Couto, B., Melloni, M., Kichic, R., Cetkovich, M., Favaloro, R., Moser, J., Manes, F., Ibanez, A., et al. (2015). The roles of interoceptive sensitivity and metacognitive interoception in panic. \emph{Behavioral and Brain Functions}, \emph{11}(1), 1--6.

>>>>

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of (fake) lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal (Aylward et al., 2017). Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond (Arnold et al., 2019; Hanania \& Smith, 2010). While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors (e.g., presence or absence of a stimuli), independent of whether one is lying or telling the truth. By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research (Oviatt et al., 2018; Rosky, 2013).

Furthermore, our results suggest that higher ToM abilities were related to slower and less confident lies, but only in the polygraph condition. While previous bodies of work have reported mixed findings regarding the association between interoception and ToM (Chiou \& Lee, 2013; Gendolla \& Wicklund, 2009; Scaffidi Abbate et al., 2016; Wundrack \& Specht, 2023), our results suggest the two are negatively linked. One possible interpretation of our findings is that people with stronger ToM abilities by default rely more on their social skills and altercentric inference when lying (i.e., they focus on - and try to read - the other person). When that mechanism is unavailable or unsuited (e.g., when there is no person to lie to - but a ``machine'' in our case), their corresponding lying ability decreases. However, in light of the current field of mixed findings relating interoception and ToM (Canino et al., 2022; Gao et al., 2019; Miller, 2015; Shah et al., 2017), future studies are necessary to investigate the interaction of these mechanisms in different social contexts.

We also found that interoceptive abilities (as indicated by the composite interoception scores) are correlated with a higher confidence in one's lies in the polygraph condition, a condition in which the attention towards internal reactions is fostered. Indeed, this is in line with previous studies that found individuals with low interoception were more averse to risk when reputational stakes were high, telling fewer egoistical lies (Vabba et al., 2022). In fact, Vabba et al. (2022) further reported that people with high interoception abilities were less likely to differ in risk-taking tendencies, telling the same number of lies regardless of the social stakes. Consistent with our results, Mohr et al. (2023) found that individuals with high interoceptive accuracy were more likely to make egocentric decisions. However, in contrast to previous studies (Füstös et al., 2013; Owens et al., 2018; Pinna \& Edwards, 2020; Pollatos et al., 2007), we did not find any significant relationship between individuals' interoception scores and their heart rate changes during their answers. This points toward a predominantly meta-cognitive effect without necessarily an actual bodily regulation (i.e., participants with good interoception feel that their lies are more convincing, but do not actively attenuate their bodily reactions).

@@

\leavevmode\vadjust pre{\hypertarget{ref-arnold2019interoception}{}}%
Arnold, A. J., Winkielman, P., \& Dobkins, K. (2019). Interoception and social connection. \emph{Frontiers in Psychology}, \emph{10}, 2589.

\leavevmode\vadjust pre{\hypertarget{ref-aylward2017impact}{}}%
Aylward, J., Valton, V., Goer, F., Mkrtchian, A., Lally, N., Peters, S., Limbachya, T., \& Robinson, O. J. (2017). The impact of induced anxiety on affective response inhibition. \emph{Royal Society Open Science}, \emph{4}(6), 170084.

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-hanania2010selective}{}}%
Hanania, R., \& Smith, L. B. (2010). Selective attention and attention switching: Towards a unified developmental approach. \emph{Developmental Science}, \emph{13}(4), 622--635.

@@

@@
